---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# alignphone

<!-- badges: start -->
<!-- badges: end -->

The goal of alignphone is to ...

## Installation

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tjmahr/alignphone")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(alignphone)

a <- wiscbet_to_ipa(
  "dh", "4", " ", 
  "l", "I", ".", "t", "4", "l", " ", 
  "b", "cI", " ", 
  "r", "ae", "n", " ",  
  "h", "oU", "m"
) |> 
  alignment_utterance("the little boy went home")
a

b <- wiscbet_to_ipa(
  "dh", "4", " ", 
  "b", "cI", " ", 
  "w", "E", "n", "t", " ", 
  "h", "oU", "m"
)|> 
  alignment_utterance("the boy went home")
b

```

By default, the alignment only rewards exact matches. These appear in the
alignment as `|`.

```{r}
ab1 <- align_phones(a, b)
ab1

ab1$scores
```

A custom comparison function can be used for alignment. This package provides
`phone_match_partial()` which assigns partial credit based similar phonetic
features. Partial matches appear in the alignment as `:`. The `as.data.frame()`
method for the alignment shows the alignment with scores.

```{r}
ab2 <- align_phones(a, b, fun_match = phone_match_partial)
ab2

as.data.frame(ab2)
```


Our current (non-R) alignment program aligns "baby sock" (child) and "we restock"
(listener) as:

```
b-eI-b-i s-------@-k
       ||        | |
w------i r-E-s-t-@-k
```

This says that the listener heard deletions of "eI" and "b" and a substitution of
[r] for /s/.

But the partial-credit implementation here yields a better alignment.

```{r}
a <- wiscbet_to_ipa("b", "eI", "b", "i", "s", "@", "k") |> 
  alignment_utterance("baby sock")
b <- wiscbet_to_ipa("w", "i", "r", "I", "s", "t", "@", "k") |> 
  alignment_utterance("we restock")
ab2 <- align_phones(a, b, phone_match_partial)
ab2
as.data.frame(ab2)
```

Here /s/ is matched with [s] and no sounds besides /t/ were deleted.


## More tests

### pretty-buddy

This is an important benchmark. Given "pretty" for *buddy*, what sound
should align at the /b/"? The default approach pairs [b] and [r]. A partial
credit scheme that assigns -.6 to mismatched vowels and -.4 to mismatched
consonants will pair [b] and [r]. Our partial credit function assigns -.2 to
mismatches that are very similar. In this case, it aligns [b] and [p].

```{r}
buddy <- str_split_at_hyphens("b-^-d-i") |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("buddy")
pretty <- str_split_at_hyphens("p-r-I-t-i") |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("pretty")

align_phones(buddy, pretty) |> 
  print() |> 
  as.data.frame()

align_phones(buddy, pretty, phone_match_partial) |> 
  print() |> 
  as.data.frame()
```

### Deleted words

This is okay. 

```{r}
point_to_teddy <- clean_old_alignment_result("p-cI-n-t t-u t-E-d-i") |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("point to teddy")
point_teddy <- clean_old_alignment_result("p-cI-n-t ----t-E-d-i") |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("point teddy")

align_phones(point_to_teddy, point_teddy, phone_match_partial)
```

This treats *hotdogs* and *hot dogs* as a mismatch on the space. It shouldn't.

```{r}
a <- "dh-oU-z i-t dh-oU-z h-@-t-d-c-g-z s-u-n" |> 
  clean_old_alignment_result() |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("those eat those hotdogs soon")
a

b <- "--------------------h-@-t  d-c-g-z------" |> 
  clean_old_alignment_result() |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("hot dogs")
b

r1 <- align_phones(a, b, phone_match_partial)
r1
```

This one works when there are word gap characters.

```{r}
he_wants_somebody_to_push_him <- clean_old_alignment_result(
  "h-i w-^-n-t-s s-^-m-b-^-d-i t-u p-U-sh h-I-m"
) |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("he wants somebody to push him")  

he_wants_to_push_him <- clean_old_alignment_result(
  "h-i w-^-n-t-s --------------t-u p-U-sh h-I-m"
) |> 
  wiscbet_to_ipa() |> 
  alignment_utterance("he wants to push him")  

r1 <- align_phones(
  he_wants_somebody_to_push_him, 
  he_wants_to_push_him, 
  phone_match_partial
)

r1
```

But when the gaps are removed, it does not work.

```{r}
remove_gaps <- function(x) {
  gaps <- x$phones %in% c(" ", ".")
  x$phones <- x$phones[!gaps]
  x
}


r1 <- align_phones(
  he_wants_somebody_to_push_him |> remove_gaps(), 
  he_wants_to_push_him |> remove_gaps(), 
  phone_match_partial
)
r1
```


### the swift benchmark

I really want this one to match `b`-`v` and `li`-`lI`, but I have to ignore word
boundaries entirely. I need to figure out how to favor the kind of alignment.

```{r}
x <- wiscbet_to_ipa(
    "l", "@", "ng", 
    "l", "I", "s", "t", 
    "^", "v", 
    "E", "k", "s", 
    "l", "^", "v", "4^", "z"
  )

y <- wiscbet_to_ipa(
    "l", "oU", "n", "l", "i", 
    "s", "t", "@", "r", "b", "^", "k", "s", 
    "l", "^", "v", "4^", "z"
  )


x <- align_phones(x, y, phone_match_partial) |> 
  set_utterance_labels(
    "long list of ex-lovers",
    "lonely starbucks lovers"
  ) |> 
  print() |> 
  as.data.frame()



```

