---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# alignphone

<!-- badges: start -->
<!-- badges: end -->

The goal of alignphone is to ...

## Installation

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tjmahr/alignphone")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(alignphone)

# the little boy ran home
a <- c(
  "ð", "ə", ".", 
  "l", "ɪ", "t", "l", ".", 
  "b", "ɔɪ", ".", 
  "r", "æ", "n", ".",  
  "h", "oʊ", "m"
)

# the boy went home
b <- c(
  "ð", "ə", ".", 
  "b", "ɔɪ", ".", 
  "w", "E", "n", "t", ".", 
  "h", "oʊ", "m"
)
```

By default, the alignment only rewards exact matches. These appear in the
alignment as `|`.

```{r}
ab1 <- align_phones(a, b)
ab1

ab1$scores
```

A custom comparison function can be used for alignment. This package provides
`phone_match_partial()` which assigns partial credit based similar phonetic
features. Partial matches appear in the alignment as `:`.

```{r}
ab2 <- align_phones(a, b, fun_match = phone_match_partial)
ab2

ab2$scores
```
